---
title: "Lab 3 - Tessellations, Point-in-Polygon"
format: 
  html: 
    code-fold: true
    code-summary: "Code"
editor: 
  markdown: 
    wrap: 72
---

Libraries:

```{r}
#| warning: FALSE
library(tidyverse)
library(sf)
library(AOI)
library(mapview)
library(rmapshaper)
library(knitr)
library(kableExtra)
library(rmarkdown)
library(ggplot2)
library(units)
library(gghighlight)
library(leaflet)
library(leafem)
library(glue)
```

## Question 1:

Here we will prepare five tessellated surfaces from CONUS and write a
function to plot them in a descriptive way.

### Step 1.1

First, we need a spatial file of CONUS counties. For future area
calculations we want these in an equal area projection (EPSG:5070).

To achieve this:

-   get an sf object of US counties (AOI::aoi_get(state = "conus",
    county = "all"))

-   transform the data to EPSG:5070

```{r}
#| warning: FALSE
conus <- AOI::aoi_get(state = "conus", county = "all") %>%
  st_transform(crs = 5070)
```

### Step 1.2

For triangle based tessellations we need point locations to serve as our
“anchors”.

To achieve this:

-   generate county centroids using st_centroid

-   Since, we can only tessellate over a feature we need to combine or
    union the resulting 3,108 POINT features into a single MULTIPOINT
    feature

-   Since these are point objects, the difference between union/combine
    is mute

```{r}
#| warning: FALSE
centroid <- conus %>%
  st_centroid() %>%
  st_combine()
```

### Step 1.3

Make a voronoi tessellation over your county centroids (MULTIPOINT)

```{r}
#| warning: FALSE
voronoi <- st_voronoi(centroid, envelope = st_union(conus)) %>%
  st_collection_extract("POLYGON") %>%
  st_as_sf() %>%
  mutate(id = row_number()) %>%
  st_cast()
```

Make a triangulated tessellation over your county centroids (MULTIPOINT)

```{r}
#| warning: FALSE
triangulated <- st_triangulate(centroid) %>%
  st_collection_extract("POLYGON") %>%
  st_as_sf() %>%
  mutate(id = row_number()) %>%
  st_cast()
```

Make a gridded coverage with n = 70, over your counties object

```{r}
#| warning: FALSE
grid <- st_make_grid(conus, n = 70, what = "polygons", square = TRUE) %>%
  st_as_sf() %>%
  mutate(id = row_number ()) %>%
  st_cast()
```

Make a hexagonal coverage with n = 70, over your counties object In
addition to creating these 4 coverage’s we need to add an ID to each
tile.

```{r}
#| warning: FALSE
hex <- st_make_grid(conus, n = 70, what = "polygons", square = FALSE) %>%
  st_as_sf() %>%
  mutate(id = row_number ()) %>%
  st_cast()
```

### Step 1.4

If you plot the above tessellations you’ll see the triangulated surfaces
produce regions far beyond the boundaries of CONUS.

We need to cut these boundaries to CONUS border.

To do this, we will call on st_intersection, but will first need a
geometry of CONUS to serve as our differencing feature. We can get this
by unioning our existing county boundaries.

```{r}
#| warning: FALSE
union <- st_union(conus) 

voronoi_union <- st_intersection(voronoi, union)

triangulated_union <- st_intersection(triangulated, union)

grid_union <- st_intersection(grid, union)

hex_union <- st_intersection(hex, union)
```

### Step 1.5

With a single feature boundary, we must carefully consider the
complexity of the geometry. Remember, the more points our geometry
contains, the more computations needed for spatial predicates our
differencing. For a task like ours, we do not need a finely resolved
coastal boarder.

To achieve this:

-   Simplify your unioned border using the Visvalingam algorithm
    provided by rmapshaper::ms_simplify.

-   Choose what percentage of vertices to retain using the keep argument
    and work to find the highest number that provides a shape you are
    comfortable with for the analysis:

```{r}
#| warning: FALSE
simple <- ms_simplify(
  union,
  keep = .05, 
  method = "vis",
  weighting = 0.7,
  keep_shapes = FALSE,
  no_repair = FALSE,
  snap = TRUE,
  explode = FALSE,
  drop_null_geometries = TRUE,
  snap_interval = NULL)
```

-   Once you are happy with your simplification, use the mapview::npts
    function to report the number of points in your original object, and
    the number of points in your simplified object.

```{r}
#| warning: FALSE
mapview::npts(union)
mapview::npts(simple)
```

-   How many points were you able to remove? What are the consequences
    of doing this computationally?

    -   10,715 points. This will make computation faster but we will
        lose detail.

-   Finally, use your simplified object to crop the two triangulated
    tessellations with st_intersection:

```{r}
#| warning: FALSE
voronoi_simple <- st_intersection(voronoi, simple)
triangulated_simple <- st_intersection(triangulated, simple)
```

### Step 1.6

The last step is to plot your tessellations. We don’t want to write out
5 ggplots (or mindlessly copy and paste 😄)

Instead, lets make a function that takes an sf object as arg1 and a
character string as arg2 and returns a ggplot object showing arg1 titled
with arg2.

For this function:

-   The name can be anything you chose, arg1 should take an sf object,
    and arg2 should take a character string that will title the plot

-   In your function, the code should follow our standard ggplot
    practice where your data is arg1, and your title is arg2

-   The function should also enforce the following:

    -   a white fill

    -   a navy border

    -   a size of 0.2

    -   \`theme_void\`\`

    -   a caption that reports the number of features in arg1

-   You will need to paste character stings and variables together.

```{r}
#| warning: FALSE
tess_plot = function(object, title_text) {
  ggplot(data = object) +
  geom_sf(fill = "white", color = "navy", size = 0.2) +
  theme_void() +
  labs(
    title = title_text,
    caption = paste("number of features:", nrow(object)))
}

tess_plot(voronoi_simple, "voronoi plot")
```

### Step 1.7

Use your new function to plot each of your tessellated surfaces and the
original county data (5 plots in total):

```{r}
#| warning: FALSE
tess_plot(conus, "CONUS County Boundaries")

tess_plot(voronoi_simple, "Voronoi Tessellation")

tess_plot(triangulated_simple, "Triangulated Tessellation")

tess_plot(grid_union, "Square Grid Tessellation")

tess_plot(hex_union, "Hexagonal Grid Tessellation")
```

## Question 2

In this question, we will write out a function to summarize our
tessellated surfaces. Most of this should have been done in your daily
assignments.

### Step 2.1

First, we need a function that takes a sf object and a character string
and returns a data.frame.

For this function:

-   The function name can be anything you chose, arg1 should take an sf
    object, and arg2 should take a character string describing the
    object

-   In your function, calculate the area of arg1; convert the units to
    km2; and then drop the units

-   Next, create a data.frame containing the following:

    -   text from arg2

    -   the number of features in arg1

    -   the mean area of the features in arg1 (km2)

    -   the standard deviation of the features in arg1

    -   the total area (km2) of arg1

-   Return this data.frame

```{r}
#| warning: FALSE
tess_sum <- function(sf_object, character_string) {
  
  area_m2 <- st_area(sf_object)
  
  area_km2 <- set_units(area_m2, "km^2") %>%
    as.numeric()
  
  data.frame(
    description = character_string,
    num_features = length(area_km2),
    mean_area_km2 = mean(area_km2),
    sd_area_km2 = sd(area_km2),
    total_area_km2 = sum(area_km2)
  )
}
```

### Step 2.2

Use your new function to summarize each of your tessellations and the
original counties.

```{r}
#| warning: FALSE
conus_sum <- tess_sum(conus, "CONUS Summary")
conus_sum

voronoi_sum <- tess_sum(voronoi_simple, "Voronoi Tessallation Summary")
voronoi_sum

triangulated_sum <- tess_sum(triangulated_simple, "Triangulated Tessellation Summary")
triangulated_sum

grid_sum <- tess_sum(grid_union, "Square Grid Tessellation Summary")
grid_sum

hex_sum <- tess_sum(hex_union, "Hexagonal Grid Tessellation Summary")
hex_sum
```

### Step 2.3

Multiple data.frame objects can bound row-wise with bind_rows into a
single data.frame

```{r}
#| warning: FALSE
sum_all <- bind_rows(
  conus_sum,
  voronoi_sum,
  triangulated_sum,
  grid_sum,
  hex_sum
)

sum_all <- sum_all %>%
  rename(
    Description = description,
    `Number of Features` = num_features,
    `Mean Area (km²)` = mean_area_km2,
    `SD Area (km²)` = sd_area_km2,
    `Total Area (km²)` = total_area_km2
  )
```

### Step 2.4

Once your 5 summaries are bound (2 tessellations, 2 coverage’s, and the
raw counties) print the data.frame as a nice table using
knitr/kableExtra.

```{r}
#| warning: FALSE
sum_all %>%
  kable(format = "html", caption = "Comparison of Spatial Coverages and Tessellations", digits = 2) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) 
```

Step 2.5 Comment on the traits of each tessellation. Be specific about
how these traits might impact the results of a point-in-polygon analysis
in the contexts of the modifiable areal unit problem and with respect
computational requirements.

Voronoi and triangulated tessellations are sensitive to the Modifiable
Areal Unit Problem (MAUP), as the shape and size of their polygons
depend on centroid placement, potentially misaligning with natural
boundaries. Grid tessellations are less affected by MAUP but may
misalign with natural features, while hexagonal tessellations offer a
more uniform and less biased representation. Computationally, grid and
hexagonal tessellations are more efficient than Voronoi and triangulated
tessellations, which require more complex geometric calculations.

## Question 3:

The data we are going to analysis in this lab is from US Army Corp of
Engineers National Dam Inventory (NID). This dataset documents \~91,000
dams in the United States and a variety of attribute information
including design specifications, risk level, age, and purpose.

For the remainder of this lab we will analysis the distributions of
these dams (Q3) and their purpose (Q4) through using a point-in-polygon
analysis.

### Step 3.1

In the tradition of this class - and true to data science/GIS work - you
need to find, download, and manage raw data. While the raw NID data is
no longer easy to get with the transition of the USACE services to ESRI
Features Services, I staged the data in the resources directory of this
class. To get it, navigate to that location and download the raw file
into you lab data directory.

-   Return to your RStudio Project and read the data in using the
    readr::read_csv

-   After reading the data in, be sure to remove rows that don’t have
    location values (!is.na())

-   Convert the data.frame to a sf object by defining the coordinates
    and CRS

-   Transform the data to a CONUS AEA (EPSG:5070) projection - matching
    your tessellation

-   Filter to include only those within your CONUS boundary

```{r}
#| warning: FALSE
dams = readr::read_csv('data/NID2019_U.csv') 

usa <- AOI::aoi_get(state = "conus") %>% 
  st_union() %>% 
  st_transform(5070)

dams2 = dams %>% 
  filter(!is.na(LATITUDE) ) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4236) %>% 
  st_transform(5070) %>% 
  st_filter(usa)
```

## Step 3.2

Step 3.2 Following the in-class examples develop an efficient
point-in-polygon function that takes:

-   points as arg1,

-   polygons as arg2,

-   The name of the id column as arg3

The function should make use of spatial and non-spatial joins, sf
coercion and dplyr::count. The returned object should be input sf object
with a column - n - counting the number of points in each tile.

```{r}
#| warning: FALSE
point_poly <- function(points, polygons, id) {
  joined <- st_join(points, polygons[id], left = FALSE)
  
  counts <- joined %>%
    st_drop_geometry() %>%
    count(!!sym(id), name = "n")
  
  polygons %>%
    left_join(counts, by = id) %>%
    mutate(n = ifelse(is.na(n), 0, n))
}
```

### Step 3.3

Apply your point-in-polygon function to each of your five tessellated
surfaces where:

-   Your points are the dams

-   Your polygons are the respective tessellation

-   The id column is the name of the id columns you defined

```{r}
#| warning: FALSE
dams_conus <- point_poly(dams2, conus, "fip_code")
dams_voronoi <- point_poly(dams2, voronoi_simple, "id")
dams_triangulated <- point_poly(dams2, triangulated_simple, "id")
dams_grid <- point_poly(dams2, grid_union, "id")
dams_hex <- point_poly(dams2, hex_union, "id")
```

### Step 3.4

Lets continue the trend of automating our repetitive tasks through
function creation. This time make a new function that extends your
previous plotting function.

For this function:

-   The name can be anything you chose, arg1 should take an sf object,
    and arg2 should take a character string that will title the plot

-   The function should also enforce the following:

    -   the fill aesthetic is driven by the count column n

    -   the col is NA

    -   the fill is scaled to a continuous viridis color ramp

    -   theme_void

    -   a caption that reports the number of dams in arg1 (e.g. sum(n))

        -   You will need to paste character strings and variables
            together.

```{r}
#| warning: FALSE
dam_plot <- function(object, title_text) {
  object <- object %>%
    filter(st_geometry_type(.) %in% c("POLYGON", "MULTIPOLYGON")) %>%
    filter(!st_is_empty(.)) %>%
    filter(st_is_valid(.)) %>%
    filter(!is.na(n)) %>%
    mutate(n = as.numeric(n))

  ggplot(data = object) +
    geom_sf(aes(fill = n), color = NA) +
    scale_fill_viridis_c(option = "viridis", na.value = "white") +
    theme_void() +
    labs(
      title = title_text,
      caption = paste("total number of dams:", sum(object$n, na.rm = TRUE)),
      fill = "dam count"
    )
}
```

### Step 3.5

Apply your plotting function to each of the 5 tessellated surfaces with
Point-in-Polygon counts:

```{r}
#| warning: FALSE
dam_plot(dams_conus, "Dam Count by County")
dam_plot(dams_voronoi, "Dam Count by Voronoi Tessellation")
dam_plot(dams_triangulated, "Dam Count by Triangulated Tessellation")
dam_plot(dams_grid, "Dam Count by Square Grid")
dam_plot(dams_hex, "Dam Count by Hexagonal Grid")
```

### Step 3.6

Comment on the influence of the tessellated surface in the visualization
of point counts. How does this related to the MAUP problem. Moving
forward you will only use one tessellation, which will you chose and
why?

I’m choosing the hexagonal grid tessellation because it gives a more
even and balanced view of point counts, making it easier to see
concentration areas without the distortion you might get with Voronoi or
triangulated tessellations. The uniform shape helps avoid the issues of
the Modifiable Areal Unit Problem (MAUP), which can mess with analysis
when shapes vary too much. It also works well for visualizing the
concentration in the Midwest and South, and it’s more computationally
efficient for larger datasets.

## Question 4

### Step 4.1

-   Your task is to create point-in-polygon counts for at least 4 of the
    follwing dam purposes:

    -   I Irrigation

    -   H Hydroelectric

    -   C Flood Control

    -   N Navigation

    -   S Water Supply

    -   R Recreation

    -   P Fire Protection

    -   F Fish and Wildlife

    -   D Debris Control

    -   T Tailings

    -   G Grade Stabilization

    -   O Other

-   You will use grepl to filter the complete dataset to those with your
    chosen purpose

-   Remember that grepl returns a boolean if a given pattern is matched
    in a string

-   grepl is vectorized so can be used in dplyr::filter

For your analysis, choose at least four of the above codes, and describe
why you chose them. Then for each of them, create a subset of dams that
serve that purpose using dplyr::filter and grepl

-   I chose Irrigation, Hydroelectric, Flood Control, and Water Supply.
    I chose these because they are uses for dams that I am most familiar
    with.

Finally, use your point-in-polygon function to count each subset across
your elected tessellation

```{r}
#| warning: FALSE
dams_i <- dams2 %>%
  filter(grepl("I", PURPOSES))

dams_h <- dams2 %>%
  filter(grepl("H", PURPOSES))

dams_c <- dams2 %>%
  filter(grepl("C", PURPOSES))

dams_s <- dams2 %>%
  filter(grepl("S", PURPOSES))
```

```{r}
#| warning: FALSE
dams_i_hex <- point_poly(dams_i, hex_union, "id")
dams_h_hex <- point_poly(dams_h, hex_union, "id")
dams_c_hex <- point_poly(dams_c, hex_union, "id")
dams_s_hex <- point_poly(dams_s, hex_union, "id")
```

### Step 4.2

-   Now use your plotting function from Q3 to map these counts.

-   But! you will use gghighlight to only color those tiles where the
    count (n) is greater then the (mean + 1 standard deviation) of the
    set

-   Since your plotting function returns a ggplot object already, the
    gghighlight call can be added “+” directly to the function.

-   The result of this exploration is to highlight the areas of the
    country with the most

```{r}
#| warning: FALSE
mean_sd_i <- mean(dams_i_hex$n) + sd(dams_i_hex$n)
mean_sd_h <- mean(dams_h_hex$n) + sd(dams_h_hex$n)
mean_sd_c <- mean(dams_c_hex$n) + sd(dams_c_hex$n)
mean_sd_s <- mean(dams_s_hex$n) + sd(dams_s_hex$n)

dam_plot(dams_i_hex, "Irrigation Dams") +
  gghighlight(n > mean_sd_i, label_key = n)

dam_plot(dams_h_hex, "Hydroelectric Dams") +
  gghighlight(n > mean_sd_h, label_key = n)

dam_plot(dams_c_hex, "Flood Control Dams") +
  gghighlight(n > mean_sd_c, label_key = n)

dam_plot(dams_s_hex, "Water Supply Dams") +
  gghighlight(n > mean_sd_s, label_key = n)
```

### Step 4.3

Comment of geographic distribution of dams you found. Does it make
sense? How might the tessellation you chose impact your findings? How
does the distribution of dams coincide with other geographic factors
such as river systems, climate, ect?

## Question 5: 

You have also been asked to identify the largest, at risk, flood control
dams in the country

You must also map the Mississippi River System - This data is available
here - Download the shapefile and unzip it into your data directory. -
Use read_sf to import this data and filter it to only include the
Mississippi SYSTEM

To achieve this:

Create an interactive map using leaflet to show the largest
(NID_STORAGE); high-hazard (HAZARD == “H”) dam in each state

-   The markers should be drawn as opaque, circle markers, filled red
    with no border, and a radius set equal to the (NID_Storage /
    1,500,000)

-   The map tiles should be selected from any of the tile providers

-   A popup table should be added using leafem::popup and should only
    include the dam name, storage, purposes, and year completed

-   The Mississippi system should be added at a Polyline feature

```{r}
major_rivers <- read_sf("data/major_rivers/MajorRivers.shp")

miss <- major_rivers %>%
  filter(SYSTEM == "Mississippi") %>%
  st_transform(crs = 4326)
```

```{r}
H <- dams2 %>%
  mutate(STATE_CODE = substr(NIDID, 1, 2)) %>%
  filter(HAZARD == "H", grepl("C", PURPOSES)) %>%
  group_by(STATE_CODE) %>%
  slice_max(order_by = NID_STORAGE, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  st_transform(crs = 4326)
```

```{r}
H$popup <- glue::glue_data(
  H,
  "<b>{DAM_NAME}</b><br/>",
  "Storage: {format(NID_STORAGE, big.mark = ',')} acre-ft<br/>",
  "Purposes: {PURPOSES}<br/>",
  "Year Completed: {YEAR_COMPLETED}"
)


leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolylines(data = miss, color = "blue", weight = 2, opacity = 0.8) %>%
  addCircleMarkers(
    data = H,
    radius = ~NID_STORAGE / 1500000,
    color = NA,
    fillColor = "red",
    fillOpacity = 0.8,
    label = ~DAM_NAME,
    popup = ~popup
  )
```
